---
title: "Homework1"
output: html_document
date: "2026-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

```{r setup-a, message=FALSE, warning=FALSE}
# If you haven't installed these yet:
# install.packages(c("glmnet", "dplyr"))

library(glmnet)  # lasso / elastic net via glmnet
library(dplyr)   # rename(), %>%, etc.
```

```{r}

```



### 1) Set your working directory (your HW folder)

You told me your folder is:

`/Users/karolinawiesiolek/Desktop/Mailman Spring 2026/DSII/HW1/Homework_1`

Use this (copy exactly):

```{r set-wd-a}
setwd("/Users/karolinawiesiolek/Desktop/Mailman Spring 2026/DSII/HW1/Homework_1")

# sanity check: shows what R thinks the folder is
getwd()

# sanity check: shows what files R sees inside this folder
list.files()
```


**Explanation**
- `setwd(...)` tells R “this is the folder where my files live.”
- `list.files()` is your fastest way to confirm the exact filenames (spaces, underscores, “(1)”, etc.).

---

### 2) Read training + test data (robust to filename variations)

This chunk tries to find the train/test/dictionary files even if they have slightly different names (like “training (1).csv”).

```{r read-data-a}
# --- Helper function: find a file by keyword pattern ---
pick_file <- function(pattern) {
  f <- list.files(pattern = pattern, ignore.case = TRUE)
  if (length(f) == 0) stop("Could not find file matching pattern: ", pattern)
  f[1]  # take the first match
}

train_file <- pick_file("train|training")
test_file  <- pick_file("test")
dict_file  <- pick_file("dict|dictionary")

train_raw <- read.csv(train_file, stringsAsFactors = FALSE, check.names = FALSE)
test_raw  <- read.csv(test_file,  stringsAsFactors = FALSE, check.names = FALSE)

dim(train_raw)
dim(test_raw)
names(train_raw)[1:10]
```

**Explanation**
- `pick_file()` searches your folder using a regex pattern.
- `read.csv(..., check.names = FALSE)` prevents R from silently changing column names (important because your response is `"Sale price"` with a space).
- `stringsAsFactors = FALSE` keeps character columns as character (we’ll handle dummy coding ourselves).

---

### 3) Standardize the response column name

Your homework says the response column is `"Sale price"`, but sometimes it appears as `"Sale_Price"` or `"SalePrice"`. This code detects it and renames it to `SalePrice` in both train/test.

```{r response-col-a}
resp_candidates <- c("Sale price", "Sale_Price", "SalePrice", "sale_price", "Sale Price")

resp_train <- resp_candidates[resp_candidates %in% names(train_raw)][1]
resp_test  <- resp_candidates[resp_candidates %in% names(test_raw)][1]

stopifnot(!is.na(resp_train), !is.na(resp_test))

train_raw <- train_raw %>% rename(SalePrice = all_of(resp_train))
test_raw  <- test_raw  %>% rename(SalePrice = all_of(resp_test))

summary(train_raw$SalePrice)
```


**Explanation**
- We search for likely spellings of the response column.
- `stopifnot(...)` immediately stops with an error if the response column wasn’t found (better than silently running wrong code).
- Renaming to `SalePrice` makes everything consistent.

---

## 4) Build the `x` matrix and `y` vector for glmnet

`glmnet` does **not** take a data frame with factors directly. It wants:
- `x`: numeric matrix (including dummy variables for categories)
- `y`: numeric response vector

`model.matrix()` does the dummy encoding for you.


```{r design-matrix-a}
# model.matrix creates dummy variables for factors/characters automatically.
# We remove the intercept column because glmnet includes an intercept by default.
x_train <- model.matrix(SalePrice ~ ., data = train_raw)[, -1]
y_train <- train_raw$SalePrice

x_test  <- model.matrix(SalePrice ~ ., data = test_raw)[, -1]
y_test  <- test_raw$SalePrice

dim(x_train)
dim(x_test)
```


**Explanation**
- `SalePrice ~ .` means “use every other column as a predictor.”
- `[, -1]` drops the intercept column.
- Dummy encoding is crucial: categorical variables become 0/1 columns.

---

## 5) Fit LASSO with 10-fold CV and get tuning parameter

Lasso is `alpha = 1`.

```{r lasso-fit-a}
set.seed(2026)  # makes CV folds reproducible

cv_lasso <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,       # alpha=1 => LASSO
  nfolds = 10,     # 10-fold cross-validation
  standardize = TRUE  # default TRUE; keeps predictors on comparable scales
)

# Plot CV curve
plot(cv_lasso)

lambda_min <- cv_lasso$lambda.min  # lambda with minimum CV error
lambda_1se <- cv_lasso$lambda.1se  # 1SE rule lambda (more regularized)

lambda_min
lambda_1se
```


**Explanation**
- `cv.glmnet()` fits many lasso models across a sequence of λ values and evaluates them by CV.
- `lambda.min` = best CV performance (lowest mean CV error).
- `lambda.1se` = simplest model whose CV error is within **1 standard error** of the minimum (tends to select fewer predictors).

---

## 6) Test error (RMSE) at λ_min (and optionally λ_1se)

The homework asks for “test error.” In regression we usually report RMSE.

```{r lasso-testerror-a}
rmse <- function(pred, truth) sqrt(mean((pred - truth)^2))

# Predictions on test set
pred_test_min <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.min"))
pred_test_1se <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.1se"))

test_rmse_min <- rmse(pred_test_min, y_test)
test_rmse_1se <- rmse(pred_test_1se, y_test)

test_rmse_min
test_rmse_1se
```



**Explanation**
- `predict(..., s="lambda.min")` uses the model tuned by CV and predicts test outcomes.
- RMSE is in the same units as the outcome (`SalePrice`), so it’s easy to interpret.

> If your instructor expects MSE instead of RMSE, just square the RMSE or compute mean((pred-truth)^2).

---

## 7) How many predictors are included with the 1SE rule?

We count how many coefficients are non-zero at `lambda.1se`.  
We do **not** count the intercept.


```{r lasso-1se-count-a}
coef_1se <- coef(cv_lasso, s = "lambda.1se")  # sparse matrix of coefficients

# Count non-zero coefficients excluding intercept (first row)
n_predictors_1se <- sum(coef_1se[-1, 1] != 0)

n_predictors_1se
```


**Explanation**
- Lasso sets many coefficients exactly to 0.
- Non-zero coefficients correspond to “selected predictors.”
- `coef_1se[-1, ]` removes the intercept row before counting.

---

## 8) Final “report” lines for Part (a)

This prints exactly what you need to paste into your write-up.


```{r report-a}
cat("Part (a) results:\n")
cat("Selected tuning parameter (lambda.min):", lambda_min, "\n")
cat("Test RMSE at lambda.min:", test_rmse_min, "\n\n")

cat("1SE rule lambda (lambda.1se):", lambda_1se, "\n")
cat("Test RMSE at lambda.1se:", test_rmse_1se, "\n")
cat("Number of predictors (nonzero) at lambda.1se:", n_predictors_1se, "\n")
```


## QUESTION 2 
