---
title: "Homework1"
output: html_document
date: "2026-02-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

I am going to add packages as needed, allows me to keep better track of questions and solving procedures. 
```{r setup-a, message=FALSE, warning=FALSE}
library(glmnet)  
library(dplyr)  
```


### 1) Working Directory

```{r set-wd-a}
setwd("/Users/karolinawiesiolek/Desktop/Mailman Spring 2026/DSII/HW1/Homework_1")

# sanity check: shows what R thinks the folder is
getwd()

# sanity check: shows what files R sees inside this folder
list.files()
```


**Explanation**
- `setwd(...)` tells R “this is the folder where my files live.”
- `list.files()` is your fastest way to confirm the exact filenames (spaces, underscores, “(1)”, etc.).

---

### 2) Read training + test data (robust to filename variations)

This chunk tries to find the train/test/dictionary files even if they have slightly different names (like “training (1).csv”).

```{r read-data-a}
# --- Helper function: find a file by keyword pattern ---
pick_file <- function(pattern) {
  f <- list.files(pattern = pattern, ignore.case = TRUE)
  if (length(f) == 0) stop("Could not find file matching pattern: ", pattern)
  f[1]  # take the first match
}

train_file <- pick_file("train|training")
test_file  <- pick_file("test")
dict_file  <- pick_file("dict|dictionary")

train_raw <- read.csv(train_file, stringsAsFactors = FALSE, check.names = FALSE)
test_raw  <- read.csv(test_file,  stringsAsFactors = FALSE, check.names = FALSE)

dim(train_raw)
dim(test_raw)
names(train_raw)[1:10]
```

**Explanation**
- `pick_file()` searches your folder using a regex pattern.
- `read.csv(..., check.names = FALSE)` prevents R from silently changing column names (important because your response is `"Sale price"` with a space).
- `stringsAsFactors = FALSE` keeps character columns as character (we’ll handle dummy coding ourselves).

---

### 3) Standardize the response column name

Your homework says the response column is `"Sale price"`, but sometimes it appears as `"Sale_Price"` or `"SalePrice"`. This code detects it and renames it to `SalePrice` in both train/test.

```{r response-col-a}
resp_candidates <- c("Sale price", "Sale_Price", "SalePrice", "sale_price", "Sale Price")

resp_train <- resp_candidates[resp_candidates %in% names(train_raw)][1]
resp_test  <- resp_candidates[resp_candidates %in% names(test_raw)][1]

stopifnot(!is.na(resp_train), !is.na(resp_test))

train_raw <- train_raw %>% rename(SalePrice = all_of(resp_train))
test_raw  <- test_raw  %>% rename(SalePrice = all_of(resp_test))

summary(train_raw$SalePrice)
```


**Explanation**
- We search for likely spellings of the response column.
- `stopifnot(...)` immediately stops with an error if the response column wasn’t found (better than silently running wrong code).
- Renaming to `SalePrice` makes everything consistent.

---

## 4) Build the `x` matrix and `y` vector for glmnet

`glmnet` does **not** take a data frame with factors directly. It wants:
- `x`: numeric matrix (including dummy variables for categories)
- `y`: numeric response vector

`model.matrix()` does the dummy encoding for you.


```{r design-matrix-a}
# model.matrix creates dummy variables for factors/characters automatically.
# We remove the intercept column because glmnet includes an intercept by default.
x_train <- model.matrix(SalePrice ~ ., data = train_raw)[, -1]
y_train <- train_raw$SalePrice

x_test  <- model.matrix(SalePrice ~ ., data = test_raw)[, -1]
y_test  <- test_raw$SalePrice

dim(x_train)
dim(x_test)
```


**Explanation**
- `SalePrice ~ .` means “use every other column as a predictor.”
- `[, -1]` drops the intercept column.
- Dummy encoding is crucial: categorical variables become 0/1 columns.

---

## 5) Fit LASSO with 10-fold CV and get tuning parameter

Lasso is `alpha = 1`.

```{r lasso-fit-a}
set.seed(2026)  # makes CV folds reproducible

cv_lasso <- cv.glmnet(
  x = x_train,
  y = y_train,
  alpha = 1,       # alpha=1 => LASSO
  nfolds = 10,     # 10-fold cross-validation
  standardize = TRUE  # default TRUE; keeps predictors on comparable scales
)

# Plot CV curve
plot(cv_lasso)

lambda_min <- cv_lasso$lambda.min  # lambda with minimum CV error
lambda_1se <- cv_lasso$lambda.1se  # 1SE rule lambda (more regularized)

lambda_min
lambda_1se
```


**Explanation**
- `cv.glmnet()` fits many lasso models across a sequence of λ values and evaluates them by CV.
- `lambda.min` = best CV performance (lowest mean CV error).
- `lambda.1se` = simplest model whose CV error is within **1 standard error** of the minimum (tends to select fewer predictors).

---

## 6) Test error (RMSE) at λ_min (and optionally λ_1se)

The homework asks for “test error.” In regression we usually report RMSE.

```{r lasso-testerror-a}
rmse <- function(pred, truth) sqrt(mean((pred - truth)^2))

# Predictions on test set
pred_test_min <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.min"))
pred_test_1se <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.1se"))

test_rmse_min <- rmse(pred_test_min, y_test)
test_rmse_1se <- rmse(pred_test_1se, y_test)

test_rmse_min
test_rmse_1se
```



**Explanation**
- `predict(..., s="lambda.min")` uses the model tuned by CV and predicts test outcomes.
- RMSE is in the same units as the outcome (`SalePrice`), so it’s easy to interpret.

> If your instructor expects MSE instead of RMSE, just square the RMSE or compute mean((pred-truth)^2).

---

## 7) How many predictors are included with the 1SE rule?

We count how many coefficients are non-zero at `lambda.1se`.  
We do **not** count the intercept.


```{r lasso-1se-count-a}
coef_1se <- coef(cv_lasso, s = "lambda.1se")  # sparse matrix of coefficients

# Count non-zero coefficients excluding intercept (first row)
n_predictors_1se <- sum(coef_1se[-1, 1] != 0)

n_predictors_1se
```


**Explanation**
- Lasso sets many coefficients exactly to 0.
- Non-zero coefficients correspond to “selected predictors.”
- `coef_1se[-1, ]` removes the intercept row before counting.

---

## 8) Final “report” lines for Part (a)

This prints exactly what you need to paste into your write-up.


```{r report-a}
cat("Part (a) results:\n")
cat("Selected tuning parameter (lambda.min):", lambda_min, "\n")
cat("Test RMSE at lambda.min:", test_rmse_min, "\n\n")

cat("1SE rule lambda (lambda.1se):", lambda_1se, "\n")
cat("Test RMSE at lambda.1se:", test_rmse_1se, "\n")
cat("Number of predictors (nonzero) at lambda.1se:", n_predictors_1se, "\n")
```


## QUESTION 2 


```{r setup-enet, message=FALSE, warning=FALSE}
library(glmnet)
library(dplyr)
```

```{r alpha-grid}
set.seed(2026)

alpha_grid <- seq(0, 1, by = 0.05)  # 0 = ridge, 1 = lasso

alpha_grid
```

# 4️⃣ For each alpha, run cross-validated glmnet

```{r enet-fit}
enet_fits <- lapply(alpha_grid, function(a) {
  cv.glmnet(
    x = x_train,
    y = y_train,
    alpha = a,
    nfolds = 10,
    standardize = TRUE
  )
})
```


# 5️⃣ Extract CV performance for each alpha


```{r enet-summary}
enet_summary <- data.frame(
  alpha = alpha_grid,
  lambda_min = sapply(enet_fits, function(f) f$lambda.min),
  lambda_1se = sapply(enet_fits, function(f) f$lambda.1se),
  cv_error_min = sapply(enet_fits, function(f) min(f$cvm)),
  cv_se_min = sapply(enet_fits, function(f) f$cvsd[which.min(f$cvm)])
)

enet_summary[order(enet_summary$cv_error_min), ][1:5, ]
```


# 6️⃣ Select the best alpha


```{r select-best-alpha}
best_index <- which.min(enet_summary$cv_error_min)

best_alpha  <- enet_summary$alpha[best_index]
best_lambda <- enet_summary$lambda_min[best_index]

best_alpha
best_lambda
```


# 7️⃣ Compute Test Error


```{r enet-test-error}
best_enet_model <- enet_fits[[best_index]]

pred_enet_test <- as.numeric(
  predict(best_enet_model, newx = x_test, s = "lambda.min")
)

rmse <- function(pred, truth) sqrt(mean((pred - truth)^2))

enet_test_rmse <- rmse(pred_enet_test, y_test)

enet_test_rmse
```



# 8️⃣ Apply 1SE rule within best alpha


```{r enet-1se}
lambda_1se_best_alpha <- best_enet_model$lambda.1se

pred_enet_test_1se <- as.numeric(
  predict(best_enet_model, newx = x_test, s = "lambda.1se")
)

enet_test_rmse_1se <- rmse(pred_enet_test_1se, y_test)

lambda_1se_best_alpha
enet_test_rmse_1se
```


# 🔎 Count predictors at 1SE


```{r enet-1se-count}
coef_1se <- coef(best_enet_model, s = "lambda.1se")
nonzero_1se <- sum(coef_1se[-1, 1] != 0)

nonzero_1se
```



---

# 🔷 Final Reporting Section (Copy for Homework)

After running, you will report:

### Selected tuning parameters:
- alpha = `best_alpha`
- lambda = `best_lambda`

### Test error:
- RMSE = `enet_test_rmse`

### 1SE Rule:
- Yes, it can be applied **within a fixed alpha**
- We selected lambda = `lambda_1se_best_alpha`
- Test RMSE at 1SE = `enet_test_rmse_1se`
- Number of predictors at 1SE = `nonzero_1se`

---

# 🧠 Conceptual Explanation (what your professor wants)

You should write something like:

> The 1SE rule is defined for a one-dimensional tuning parameter. Since elastic net has two tuning parameters (alpha and lambda), the 1SE rule cannot be applied directly across the full two-dimensional grid. However, after selecting the optimal alpha based on minimum cross-validated error, the 1SE rule can be applied to lambda within that fixed alpha. This yields a more regularized and typically more stable model.

---

# 🎯 Big Picture

- If alpha is close to 1 → data behaves like lasso
- If alpha is close to 0 → predictors are highly correlated (ridge behavior)
- Elastic net balances sparsity + grouping effect

---

If you want next:
- We can compare Lasso vs Elastic Net performance
- Or move to PLS
- Or I can help you write the exact final paragraph

You’re doing this exactly the right way.




# QUESTION 3 

```{r setup-pls, message=FALSE, warning=FALSE}
library(pls)
```



```{r pls-fit}
set.seed(2026)

pls_fit <- plsr(
  SalePrice ~ .,       # use all predictors
  data = train_raw,
  scale = TRUE,        # IMPORTANT: PLS should scale predictors
  validation = "CV"    # 10-fold cross-validation
)

summary(pls_fit)
```


---

## 🔎 What just happened?

PLS:

- Creates **latent components**
- Each component is a linear combination of predictors
- Components are chosen to:
  - explain predictor variation
  - explain response variation

Unlike lasso:
- PLS does NOT shrink coefficients to zero
- It reduces dimensionality instead

---

# 2️⃣ Plot CV error curve


```{r pls-plot}
validationplot(pls_fit, val.type = "MSEP", legendpos = "topright")
```



---

## 🔎 What this plot shows

- X-axis = number of components  
- Y-axis = CV mean squared prediction error  
- We choose the number of components that minimizes CV error  

---

# 3️⃣ Extract optimal number of components


```{r pls-select-ncomp}
cv_error <- RMSEP(pls_fit)

# RMSEP output structure:
# cv_error$val[estimate, response, components]
# First element is intercept-only model, so subtract 1

ncomp_best <- which.min(cv_error$val[1, 1, ]) - 1

ncomp_best
```


---

## 🔎 Why subtract 1?

Because:

- Index 1 = intercept-only model  
- Index 2 = 1 component  
- Index 3 = 2 components  

So we subtract 1 to get the actual number of components.

---

# ✅ This is your answer for:

> How many components are included in your model?

---

# 4️⃣ Compute Test Error

Now we evaluate the selected model on the **test data**.


```{r pls-test-error}
pred_pls_test <- as.numeric(
  predict(pls_fit, newdata = test_raw, ncomp = ncomp_best)
)

rmse <- function(pred, truth) sqrt(mean((pred - truth)^2))

pls_test_rmse <- rmse(pred_pls_test, y_test)

pls_test_rmse
```


---

# ✅ This is your answer for:

> Report the test error.

---

# 🔷 Final Reporting Section (Copy for Homework)

After running the code, report:

- Number of components selected: `ncomp_best`
- Test RMSE: `pls_test_rmse`

---

# 🧠 What This Means Conceptually

PLS differs from lasso and elastic net:

| Method | What it does |
|--------|-------------|
| Lasso | Shrinks some coefficients to exactly zero |
| Elastic Net | Shrinks + groups correlated predictors |
| PLS | Creates new orthogonal components |

PLS is especially useful when:
- Predictors are highly correlated
- There are many predictors relative to sample size

Instead of selecting variables, PLS compresses them into components.

---

# 🎯 How your professor expects you to interpret this

If:
- PLS test RMSE is lowest → PLS may generalize best  
- PLS uses many components → model is closer to OLS  
- PLS uses few components → strong dimensionality reduction  

---

If you'd like next:
- We compare Lasso vs Elastic Net vs PLS  
- Or I help you write the final paragraph for submission  

You're doing this very correctly.


# QUESTION 4 


```{r}
ls()
```

```{r rename-rmses}
lasso_test_rmse_min <- test_rmse_min
lasso_test_rmse_1se <- test_rmse_1se
```



Now Part (d) code will work exactly as written.

---

# Option B: Use your existing names directly (no renaming)



```{r enet-make-1se-rmse}
# 1) Make elastic net 1SE predictions from the best elastic net CV object
pred_enet_test_1se <- as.numeric(
  predict(best_enet_model, newx = x_test, s = "lambda.1se")
)

# 2) Compute and store the test RMSE for elastic net at lambda.1se
enet_test_rmse_1se <- rmse(pred_enet_test_1se, y_test)

enet_test_rmse_1se
```



### Why this works
- `best_enet_model` is your **selected elastic net CV model** (you already have it).
- `s = "lambda.1se"` forces the **1SE lambda** within that chosen alpha.
- We store it as `enet_test_rmse_1se` so Part (d) can use it.

---

# ✅ After that, Part (d) comparison will work

Use this (with your current lasso variable names):


```{r compare-models-d}
results <- data.frame(
  Model = c("Lasso (lambda.min)",
            "Lasso (lambda.1se)",
            "Elastic Net (lambda.min)",
            "Elastic Net (lambda.1se)",
            "PLS"),
  Test_RMSE = c(
    test_rmse_min,
    test_rmse_1se,
    enet_test_rmse,
    enet_test_rmse_1se,
    pls_test_rmse
  )
)

results[order(results$Test_RMSE), ]
```

Answer: 
The elastic net model using the 1SE rule achieved the lowest test RMSE (20516.78) among all fitted models. Therefore, it provides the best predictive performance for the housing sale price.

Elastic net combines the advantages of ridge and lasso penalties, allowing it to handle correlated predictors while still performing variable selection. Given that housing characteristics such as square footage, room counts, and quality measures are often highly correlated, elastic net is well-suited for this dataset.

Additionally, the 1SE version of the model outperformed the lambda.min version, suggesting that slightly stronger regularization improves generalization to new data. For these reasons, the elastic net model with lambda selected via the 1SE rule is chosen as the best predictive model.

# QUESTION 5 
Pakcages:
```{r e-packages, message=FALSE, warning=FALSE}
library(caret)
library(glmnet)
```


---

## 1️⃣ What we already have from (a)

From your previous work you already have:

- `cv_lasso` (from `cv.glmnet`)
- `lambda_min` and `lambda_1se` (from glmnet)
- `x_train, y_train` (design matrix + response)

Quick check:


```{r e-check}
lambda_min
lambda_1se
```



**Explanation**
- `cv.glmnet()` returns two main “chosen” lambdas:
  - `lambda.min` (min CV error)
  - `lambda.1se` (1SE rule)

---

## 2️⃣ Fit Lasso using **caret**

### Key point from lecture:
**Always provide a lambda grid** in caret so each fold uses the same lambda values.


```{r e-caret-lasso}
set.seed(2026)

ctrl <- trainControl(method = "cv", number = 10)

# Use a wide grid on a log scale (common choice)
lambda_grid <- exp(seq(log(1e-4), log(1e4), length.out = 200))

lasso_caret <- train(
  x = x_train,
  y = y_train,
  method = "glmnet",
  trControl = ctrl,
  tuneGrid = expand.grid(alpha = 1, lambda = lambda_grid)
)

lasso_caret$bestTune
```



**Explanation**
- `method="glmnet"` in caret calls `glmnet` internally.
- `tuneGrid` tells caret exactly which `alpha` and `lambda` values to consider.
- `bestTune$lambda` is the lambda caret selected (min CV RMSE).

---

## 3️⃣ Compare the selected tuning parameters


```{r e-compare-lambda}
comparison <- data.frame(
  Approach = c("glmnet: lambda.min", "glmnet: lambda.1se", "caret: bestTune lambda"),
  alpha = c(1, 1, 1),
  lambda = c(lambda_min, lambda_1se, lasso_caret$bestTune$lambda)
)

comparison
```



**What to report**
- Your “selected tuning parameter” from caret is: `lasso_caret$bestTune$lambda`
- Your “selected tuning parameters” from glmnet are: `lambda.min`, `lambda.1se`

---

## 4️⃣ (Optional but recommended) Make the comparison *fully fair*

By default, `cv.glmnet()` chooses its **own lambda sequence** unless you supply one.
Caret uses your supplied `lambda_grid`.

So to compare fairly, refit `cv.glmnet()` using the **same** lambda grid:

```{r e-glmnet-same-grid}
set.seed(2026)

cv_lasso_samegrid <- cv.glmnet(
  x = x_train, y = y_train,
  alpha = 1,
  nfolds = 10,
  lambda = lambda_grid   # <-- force the same grid used by caret
)

lambda_min_samegrid <- cv_lasso_samegrid$lambda.min
lambda_1se_samegrid <- cv_lasso_samegrid$lambda.1se

data.frame(
  Approach = c("glmnet (same grid): lambda.min", "glmnet (same grid): lambda.1se", "caret: bestTune lambda"),
  lambda = c(lambda_min_samegrid, lambda_1se_samegrid, lasso_caret$bestTune$lambda)
)
```



**Explanation**
- Now both caret and glmnet are optimizing over the **same candidate lambdas**, so differences (if any) come mainly from fold splits / metric computations.

---

## 5️⃣ Test RMSE comparison (optional, but nice to include)


```{r e-test-rmse-compare}
# caret prediction on test
pred_caret <- predict(lasso_caret, newx = x_test)
rmse_caret <- rmse(pred_caret, y_test)

# glmnet prediction on test at lambda.min and lambda.1se
pred_glmnet_min <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.min"))
pred_glmnet_1se <- as.numeric(predict(cv_lasso, newx = x_test, s = "lambda.1se"))

rmse_glmnet_min <- rmse(pred_glmnet_min, y_test)
rmse_glmnet_1se <- rmse(pred_glmnet_1se, y_test)

data.frame(
  Approach = c("caret (bestTune)", "glmnet (lambda.min)", "glmnet (lambda.1se)"),
  Test_RMSE = c(rmse_caret, rmse_glmnet_min, rmse_glmnet_1se)
)
```



---

# 🧠 Write-up: Why might caret vs glmnet choose different lambdas?

Here is a submission-ready explanation:

### Explanation paragraph (copy-paste)

Caret and glmnet can select different tuning parameters even when fitting the “same” lasso model because the selection depends on the resampling procedure and the candidate lambda values considered. First, `cv.glmnet()` generates its own lambda sequence by default (based on the data), whereas caret only searches over the lambda grid provided in `tuneGrid`; if the grids differ, the selected lambdas can differ. Second, the cross-validation folds may differ unless the exact same fold indices and seeds are enforced, which can change which lambda minimizes the CV error. Third, caret optimizes RMSE by default (depending on settings), while glmnet’s CV curve is based on mean squared error, and minor differences in how errors are aggregated and summarized can shift the chosen tuning parameter. Finally, preprocessing details (standardization, handling of missing values, and dummy-variable coding when using formula vs matrix interfaces) can cause small discrepancies. For a fair comparison, it is best to supply the same lambda grid to `cv.glmnet()` and to caret, and to use identical CV folds.

---

## What you should report in (e)

1) The lambda selected by **caret**:  
`lasso_caret$bestTune$lambda`

2) The lambda(s) selected by **glmnet**:  
`lambda_min` and `lambda_1se`

3) Whether they match; if not, use the explanation above.

---

If you paste your `comparison` table output, I’ll help you write 2–3 sentences *specific to your numbers* (i.e., “caret chose λ=..., glmnet chose λ=..., the discrepancy likely comes from ...”).
